Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-09-01 interimreport2.Rnw
minor revisions

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations. We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:
\begin{itemize}
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
\end{itemize}

@@

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago demoninator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).

@@

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age
\item Pacific people are statistically more at risk of measles infection
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
\item There is some statistical support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
\end{itemize}

@@

\begin{equation} \label{eq:pc}
p_c = 1-(1/R_0)
  \end{equation}
To acocunt for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.

We then assume that the measles epidemics are occuring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$. We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}). Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.

@@

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}). However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary. Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached. Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.

@@

The direct medical and public health costs in response to a single case of refugee-imported measles has been reported \citep{coleman12}.  Costs included labor, translation and benefits for public health workers. In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis. In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881. In addition, per-contact costs amounted to US\$264. The cost of hospitalisation for the 3-day stay by the index case was US\$931. Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents' exposed to measles (US\$240) for a total of US\$23,816.

@@

Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
\midrule
2009-2014 &	0.92-1.19	& 0.92	& 0	    & na	  & 0	      & 130	    & 22  	& 148551	  & 130519	  & 13	& 130	  & 22	& 148551	& 0	      & 0	        & 0	        & na \\
2009-2014	& 0.92-1.19	& 1.19	& 80000	& 20	  & 1600000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 126359360	& 79.97 \\
2009-2014	& 0.92-1.19	& 1.19	& 80000	& 50	  & 4000000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 123959360	& 31.99 \\
2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 336060587	& 81.72 \\
2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 329815997	& 32.69 \\
2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 377382119	& 75.71 \\
2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 369805289	& 30.28 \\

@@

\begin{sidewaystable}[htbp]  
  \centering  
  \begin{raggedright}
  \caption{Cost benefit analyses using historic epidemic sizes since 1997}  
  \label{tab:cost_ave}
   \begin{threeparttable}[b] \tiny
    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
    \toprule  
Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
\midrule

@@

\subsection{Benefit--cost analysis discussion}
The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak. The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses. However, for all analyses more detailed costs of all aspects of hopsitalisation would be beneficial. So too would information of the cost of measles vaccination in New Zealand.

We have used standard discounting from healthcare of 3\% per annum \citep{honeycutt6}, but again there may be official New Zealand discounting rates that we are unaware of.
For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 (Table ~\ref{table:direct}). Including these could be done, but we would like some further data of these before including them in our benefit--cost analsyes. It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.

@@

Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the presently naive and susceptible New Zealand population until overyone has been infected. Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}. However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimated to be more than 18 (i.e. 1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}). The benefits of catch up vaccination are clear once $R_v$ is greater than one (Table~\ref{tab:cost_sim}). If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.

Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population. The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{sec:epidemic_modelling}). Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).

\subsection{Benefit--cost analysis summary}
\begin{itemize}
\item The mean wage losses per measles case is estimated to be \$852
\item The mean cost per measles case attending hospital is estimated to be \$1710
\item Approximately 17\% of measles cases attend hopsital
\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
\end{itemize}

\section {Summary of key findings}
\begin{itemize}
\item Risk of measles infection decreases significantly with age
\item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk of measles than European or Maori children of the same age
\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, upper 95\% confidencelimit 2013--2014 outbreak, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
\item The mean wage losses per measles case is estimated to be \$852
\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hopsital
\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
\end{itemize}

>>>>

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations. We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:
\begin{itemize}
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
\end{itemize}

@@

With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).

@@

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age
\item Pacific people are statistically more at risk of measles infection
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children
\end{itemize}

@@

\begin{equation} \label{eq:pc}
p_c = 1-(1/R_0)
  \end{equation}
To account for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.

We then assume that the measles epidemics are occurring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$. We estimated the susceptible population in New Zealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}). Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.

@@

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}). However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary. Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached. Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.

@@

The direct medical and public health costs in response to a single case of refugee-imported measles has been reported \citep{coleman12}.  Costs included labour, translation and benefits for public health workers. In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis. In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881. In addition, per-contact costs amounted to US\$264. The cost of hospitalisation for the 3-day stay by the index case was US\$931. Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents' exposed to measles (US\$240) for a total of US\$23,816.

@@

Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
\midrule
2009-2014 &	0.92-1.19	& 0.92	& 0	    & na	  & 0	      & 130	    & 22  	& 148551	  & 130519	  & 13	& 130	  & 22	& 148551	& 0	      & 0	        & 0	        & na \\
2009-2014	& 0.92-1.19	& 1.19	& 80000	& 20	  & 1600000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 126359360	& 79.97 \\
2009-2014	& 0.92-1.19	& 1.19	& 80000	& 50	  & 4000000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 123959360	& 31.99 \\
2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 336060587	& 81.72 \\
2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 329815997	& 32.69 \\
2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 377382119	& 75.71 \\
2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 369805289	& 30.28 \\

@@

\begin{sidewaystable}[htbp]  
  \centering  
  \begin{raggedright}
  \caption{Cost benefit analyses using historic epidemic sizes since 1997}  
  \label{tab:cost_ave}
   \begin{threeparttable}[b] \tiny
    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
    \toprule  
Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
\midrule

@@

\subsection{Benefit--cost analysis discussion}
The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak. The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses. However, for all analyses more detailed costs of all aspects of hospitalisation would be beneficial. So too would information of the cost of measles vaccination in New Zealand.

We have used standard discounting from healthcare of 3\% per annum \citep{honeycutt6}, but again there may be official New Zealand discounting rates that we are unaware of.
For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 (Table ~\ref{table:direct}). Including these could be done, but we would like some further data of these before including them in our benefit--cost analyses. It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.

@@

Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate among the presently naive and susceptible New Zealand population until everyone has been infected. Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}. However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimated to be more than 18 (i.e. 1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}). The benefits of catch up vaccination are clear once $R_v$ is greater than one (Table~\ref{tab:cost_sim}). If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.

Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local naive population. The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{sec:epidemic_modelling}). Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).

\subsection{Benefit--cost analysis summary}
\begin{itemize}
\item The mean wage losses per measles case is estimated to be \$852
\item The mean cost per measles case attending hospital is estimated to be \$1710
\item Approximately 17\% of measles cases attend hospital
\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
\end{itemize}

\section {Summary of key findings}
\begin{itemize}
\item Risk of measles infection decreases significantly with age
\item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
\item Pacific and Asian children in the 6--17 year age categories have been at lower risk of measles than European or Maori children of the same age
\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, upper 95\% confidencelimit 2013--2014 outbreak, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
\item The mean wage losses per measles case is estimated to be \$852
\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
\end{itemize}
